Modelling the gas kinematics of an atypical Lyman alpha emitting compact dwarf galaxy

### 2017-09-01 paper/paper.tex
new discussion + new references

Star-forming Compact Dwarf Galaxies (CDGs) 
resemble the expected pristine conditions of the first galaxies in the
Universe and are the best systems to test models on primordial galaxy
formation and evolution. 
Here we report on one of such CDGs, \tol, which presents
a broad, single peaked, highly symmetric \lya emission line that had
evaded theoretical interpretation so far.  
In this paper we explain these features by two different physically
motivated kinematic models: 
an interstellar medium composed by outflowing clumps with 
random motions and an homogeneous gaseous sphere undergoing solid body
rotation.
It is the first time that an observed \lya spectrum can be explained
assuming either of these kinematic conditions.
The multiphase model requires a clump velocity dispersion of
\sigmaclump\ with outflows of \inftyclump\, while the
bulk rotation velocity is constrained to be $348^{+75}_{-48}$ \kms.
We argue that the results from the multiphase model are broadly
consistent with observations of other CDGs. 
The clump velocity dispersion implies a dynamical mass of $2\times
10^{9}$ M$_{\odot}$, ten times its baryonic mass, which can be
explained if \tol\ is hosted by a dark matter halo $6\times 10^{11}$
M$_{\odot}$. 
If future kinematic maps of \tol\ confirm the velocities suggested by
the multiphase model, it would provide additional support to expect
such kinematic state in primordial \lya emitting galaxies.   
\end{abstract}

@@

The models we use correspond to simplified geometrical configurations.
This allows us to perform a deep exploration of parameter space and
gain some physical insight into \tol's kinematic properties.
In this paper we show, for the first time in the literature, that \tol's
\lya profile can be explicitly modeled by either of these two
models.

@@

Which model could be closer to the actual kinematic conditions in \tol?  
Recent Integral Field Unit (IFU) observation of other CDGs seem to
favor the multiphse model \citep{2015A&A...577A..21C,2017A&A...600A.125C}. 
These observations  were performed with the Visible Multi-Object Spectrograph (VIMOS)
\citep{2003SPIE.4841.1670L}.
The spatial smapling was $0.67$ arcseconds and covered about
$30^{\prime\prime}\times 30^{\prime\prime}$ on the sky. 
They observed nine BCGs to produce two-dimensional maps of the
continuum and promiment emission lines, this allowed them to produce 
two-dimensional velocitity field maps using the H$\alpha$ emission
line.
They find velocity fields ranging from simple rotation patterns 
(with amplitudes of $10-120\kms$) to highly irregular. 
The velocity dispersion in most cases is typically less than $50$ \kms
with the exception of one galaxy that shows a dispersion of $130$
\kms.

The hybrid case of a rotating multiphase configuration is also of
interest, but cannot be simulated with in current models. 
However, it is the subject of ongoing work  (Remolina-Gutierrez et
al. in prep.).

We argue that that the rotation model should be discarded by
considering that the prefered rotational velocity of $V_{m}\sim350$
\kms\ is of the same order as the largest rotational velocity of giant
disc galaxies that have two orders of magnitude more baryonic mass
than \tol\ \citep{2015MNRAS.448.1767C}.

Furthermore, \tol's structural properties seem to be similar to known dwarf
elliptical (dE) galaxies. 
The sample of 7 dE galaxies observed by
\cite{2003AJ....126.1794G} shows similar properties as \tol. 
For instance, the dE galaxies IC 3344 and IC 794 have similar absolute $M_V$
magnitudes (-17.10 and -17.19) and sizes (0.90 kpc and 0.82 kpc) as
\tol. 
These dE galaxies have velocity dispersions of $31.1$ \kms and
$44.6$ \kms, respectively,  and their rotational velocities are lower
than $1$ \kms.  
This leads us to prefer the multiphase model as it provides 
a broad consistency with kinematic properties of dE galaxies.

Following the results of \cite{2011ApJ...726..108T}, who made a
detailed analysis for all kind of elliptical galaxies including dE galaxies, we
estimate a value for the dynamical mass using the constraints for
the velocity dispersion, $\sigma$,  in a spherical system localized in a
region of size $r$

@@

Assuming that the \lya emission is entirely powered by star formation 
we use the 3D half-light radius $r_s=2.25$ kpc as the typical size
for the HI region, in the multiphase model the clump velocity dispersion
$\sigma_{\rm{cl}}=$\sigmaclump thus corresponds to a dynamical mass of
$M_{\rm dyn}=2.31\pm0.04 \times 10^{9}$ $M_{\odot}$, which is ten
times the estimated baryonic mass in \tol.

Figure \ref{fig:mass} illustrates  how such dynamical
mass can be explained if \tol\ is hosted by a dark matter halo of at
$\sim 6\times 10^{11}$ M$_{\odot}$ in virial mass. 
This estimate is based on computing the integrated mass profile of a
spherical dark matter halo with a Navarro-Frenk-White profile with its
concentration following the median mass-concentration
relationship found in the Bolshoi simulation
\citep{2012MNRAS.423.3018P,2016ApJ...832..169P}.

These results would imply that \tol\ is part of the dE
class. 
However, this hypothesis opens up two puzzles.
First, dE galaxies are found either in clusters or groups of
galaxies; while \tol\ does not belong either to a group or a cluster
\citep{2001AJ....121..169F}. 
The second issue is that, unlike a dE, \tol\ hosts a bright and compact
starburst; imaging and spectroscopy studies suggest that the
starburst started less than 4 Myr ago, while the low surface
brightness regins are consistent with an age below one Gyr
\citep{2001AJ....121..169F}.

We suggest that new observational tests are needed to clarify the
kinematic nature of \tol.
Integral field unit measurements resolving its spatial extent are up
to the task.  
\tol\ spans a region of $4$ arcseconds,
an instrument such as the Multi Unit Spectroscopic Explorer (MUSE)
\citep{2014Msngr.157...13B} with its nominal $0.2$ arcseconds spatial
sampling over a $1.0$ arcminute field in wide-field mode could provide
a coarse mapping of different ionization lines to infer a kinematic
map in the H$\alpha$ line, following the kind of work presented by
\citet{Herenz16} done with the Potsdam Multi-Aperture Spectrometer
(PMAS) \citep{PMAS} on nearby \lya emitting galaxies.

Another observational hint to clarify \tol's state could be
the measurement of the \lya ionizing continuum escape fraction. 
The multiphase model predicts that averaging over all sightlines
it should be around $0.5^{+1.0}_{-0.4}$\%, with the possibility of strong
variations depending on viewing angle, i.e. some sightlines can have
an escape fraction close to $100\%$ \citep{Gronke2016}.

@@

In this particular case, we prefer the multiphase model because it has kinematic
conditions similar to those in dwarf ellipticals, while the
rotational model produces rotational velocities too high for a dwarf
galaxy.

However, \tol's environment and star forming activity are radically
different from those of dwarf ellipticals.
We suggest that additional integral field unit observations are the
only way to be certain about the detailed kinematic structure in \tol.

>>>>

Star-forming Compact Dwarf Galaxies (CDGs) 
resemble the expected pristine conditions of the first galaxies in the
Universe and are the best systems to test models on primordial galaxy
formation and evolution. 
Here we report on one of such CDGs, \tol, which presents
a broad, single peaked, highly symmetric \lya emission line that had
evaded theoretical interpretation so far.  
In this paper we explain these features by two different physically
motivated kinematic models: 
an interstellar medium composed by outflowing clumps with 
random motions and an homogeneous gaseous sphere undergoing solid body
rotation.
It is the first time that an observed \lya spectrum can be explained
assuming either of these kinematic conditions.
The multiphase model requires a clump velocity dispersion of
\sigmaclump\ with outflows of \inftyclump\, while the
bulk rotation velocity is constrained to be $348^{+75}_{-48}$ \kms.
We argue that the results from the multiphase model are broadly
consistent with observations of other CDGs. 
If \tol\ is dispersion dominated, the clump velocity dispersion
implies a dynamical mass of $2\times 10^{9}$ M$_{\odot}$, ten times
its baryonic mass, which can be explained if \tol\ is hosted by a dark
matter halo $6\times 10^{11}$ M$_{\odot}$. 
If future kinematic maps of \tol\ confirm the velocities suggested by
the multiphase model, it would provide additional support to expect
such kinematic state in primordial \lya emitting galaxies.   
\end{abstract}

@@

Observations of other Compact Dwarf Galaxies (CDGs) also show 
that the kinematics can renge from pure rotation and low velocity
dispersion to high velocity dispersion without rotaion.
\citep{2015A&A...577A..21C,2017A&A...600A.125C}.

The models we use correspond to simplified geometrical configurations.
This allows us to perform a deep exploration of parameter space and
gain some physical insight into \tol's kinematic properties.
In this paper we show, for the first time in the literature, that \tol's
\lya profile can be explicitly modeled by either of these two
models. 
The hybrid case of a rotating multiphase configuration
(as observed in some of the galaxies in \cite{2015A&A...577A..21C}) is
the subject of ongoing research (Remolina-Gutierrez et al. in prep.).  as
it cannot be simulated with existing models.

@@

Which model could be closer to the actual kinematic conditions in \tol?  
Recent Integral Field Unit (IFU) observations of other CDGs seem to
favor the multiphase model \citep{2015A&A...577A..21C,2017A&A...600A.125C}. 
These observations  were performed with the Visible Multi-Object Spectrograph (VIMOS)
\citep{2003SPIE.4841.1670L}.
The spatial smapling was $0.67^{\prime\prime}$ and covered about
$30^{\prime\prime}\times 30^{\prime\prime}$ on the sky. 
They observed nine BCGs to produce two-dimensional maps of the
continuum and promiment emission lines, this allowed them to produce 
two-dimensional velocitity field maps using the H$\alpha$ emission
line.
They find velocity fields ranging from simple rotation patterns 
(with amplitudes of $10-120$ \kms) to highly irregular. 
The typical velocity dispersion values are in the range $10-50$ \kms
with the exception of one galaxy that shows a dispersion of $130$
\kms.

These results disfavor the high rotational velocity of $V_{m}\sim350$
\kms\  that we find in the pure rotation model. 
On the other hand the results from the multiphase model yield a
velocity dispersion $\sigma=$\sigmaclump, quite close to observed
values in other CDG, favoring the multiphase model.

In this case we can  estimate a value for the dynamical
mass using the constraints for the velocity dispersion, $\sigma$,  in
a spherical system localized in a region of size $r$

@@

Assuming that the \lya emission is entirely powered by star formation 
we use the 3D half-light radius $r_s=2.25$ kpc as the typical size
for the HI region, in the multiphase model the clump velocity dispersion
$\sigma_{\rm{cl}}=$\sigmaclump\  corresponds to a dynamical mass of
$M_{\rm dyn}=2.31\pm0.04 \times 10^{9}$ $M_{\odot}$, which is ten
times the estimated baryonic mass in \tol.
This would mean that \tol\ is dark matter dominated.

Following the methodology of \citet{2011ApJ...726..108T} we estimate 
that a dark matter halo of at $\sim 6\times 10^{11}$ M$_{\odot}$ in
virial mass can explain this dynamical mass.
This estimate is based on computing the integrated mass profile of a
spherical dark matter halo with a Navarro-Frenk-White profile with its
concentration following the median mass-concentration
relationship found in the Bolshoi simulation
\citep{2012MNRAS.423.3018P,2016ApJ...832..169P}.  
Figure \ref{fig:mass} shows the enclosed mass as a function of radius
for different dark matter halos and the estimates for \tol.

New IFU observations are needed to completeley confirm \tol's.
This could be done with the Multi Unit Spectroscopic Explorer (MUSE) 
\citep{2014Msngr.157...13B}, the Gemini Multi-Object Spectrographs (GMOS) 
\citep{2004PASP..116..425H} or the Fibre Large Array Multi Element
Spectrograph (\citep{2002Msngr.110....1P})

@@

In this particular case, we prefer the multiphase model because it has kinematic
conditions similar to those in similar compact dwarf galaxies observed
with integral field unit spectroscopy, while the
rotational model produces rotational velocities too high for a dwarf
galaxy.  
New IFU observations are the only way to be certain about the detailed
kinematic structure in \tol.
